The effect of inflammation, SARS-CoV-2 infection, age and mental health on serotonin, and kynurenine and catecholamine pathway metabolites

### 2023-06-11 paper/markdown/rebuttal_letter.Rmd

---
title: "The effect of inflammation, SARS-CoV-2 infection, age and mental health on serotonin, and kynurenine and catecholamine pathway metabolites"
subtitle: "Rebuttal Letter"
author: "PSY Team"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

@@

output: 
  bookdown::word_document2:   
    reference_docx: ms_template.docx

bibliography: infl_biblio.bib
csl: psychoneuroendocrinology.csl
header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

@@

\newpage

# Editor

<span custom-style = 'reviewer'>Thank you for the timely and well written piece. 
I have received three reviews from exceptional scholars in the field. 
As you can see, there are many strengths to the paper that they note and the publication is timely. 
However, there are also several concerns raised and I am not entirely sure they can all be sufficiently addressed to warrant publication. 
However, I think it is worth a shot. 
Please be attentive to reviewer concerns for a resubmission.</span>

Thank you for appreciation of our manuscript and giving us a chance to address reviewers' concerns. 
First we would like to thank the reviewers for their very careful, thoughtful and helpful assessment and comments. 
We acknowledge that such detailed reviews take a lot of time and demand a lot of commitment and hope that we have done them justice in the revised manuscript.

Overall, we agree that the initial manuscript version was not sufficiently coherent and clear likely because of the involvement of a large team of researchers in the study. 
Following your advice and the reviewers' suggestions, we reworked thoroughly the analysis and the text. 
We would like to point out that this is one of very few manuscripts characterizing activity of the kynurenine and catecholamine pathways in COVID-19 [@Santiago-Mujika2022; @Saito2022; @Lionetto2021; @Bizjak2022; @Dewulf2022; @Ceballos2022]. 
Importantly, their role in symptoms of mental health disorders in COVID-19 patients has been proposed [@Kucukkarapinar2022; @Bower2022]. 
By analyzing two cohorts, we were able to identify key factors affecting the activity of those metabolic pathways and investigate their temporal activity profile in acute and sub-acute SARS-CoV-infection and recovery, which adds to novelty of our work.

We have re-analysed the data to address the reviewers´ concerns and have rewritten many parts of the manuscript following the reviewers' suggestions while keeping within the word limits of the journal. 
Despite extensive changes on the manuscript text, the main results remain unaltered.
In particular, we now clearly and explicitly state the gap in the literature in __Introduction__ and the analysis endpoints in the __Methods__ and __Results__ sections. 
To facilitate interpretation of our results, we now follow a top - bottom analysis approach. 
In brief, we employed the most robust and comprehensive analytic technique, multi-parameter modeling, to identify the most relevant explanatory factors affecting the key metabolites and activity markers of the kynurenine pathway, serotonin availability and the catecholamine pathway. 
Subsequently, we explore these factors in more detail by correlation and network analysis, and canonical statistical tests. 
Following the reviewer's suggestion we also keep analysis results in the SIMMUN and INCOV cohort separately for the sake of clarity. 
We have re-arrange and rewritten the Discussion section taking into account the reviewer´s suggestions.

Please see below the point by point response for details. 
Please notice that due to the scale of changes in the text, it was not feasible to mark the changes introduced during revision. 
To meet the journals' requirements concerning the maximal number of tables and figures, we had to eliminate the original figure 1 (model of investigated pathways) and move the scheme of analysis inclusion and strategy flowchart to __Supplementary Material__ (__Supplementary Figure S1__). 
Since this improved flowchart is very helpful for the understanding of our analysis, we would move it back to the main text if we are granted to exceed the maximum figure number upon acceptance.
We hope that our revised manuscript is now suitable for publication in _Psychoneuroendocrinology_.

@@

# Reviewer 1

## Issue 1

<span custom-style = 'reviewer'>The study assesses the relationship between psychological variables, inflammatory markers, and concentrations of aminoacids in individuals after SARS-CoV-2 infection compared to uninfected controls in two distinct cohorts (SIMMUN; INCOV). 
Moreover, the difference between the control group and different states after SARS-CoV-2 infection (acute, sub-acute, recovery) are compared to the control group within the longitudinal INCOV study cohort.
In addition, the study aims to test the hypothesis that Phenylalanine, Tyrosine, Tryptophan, Kynurenine, and their respective ratios, are predicted by psychological variables inflammation, and SARS-CoV-2 infection-related variables (infection yes vs no, Anti-Antibody titers).
First, it should be mentioned that the graphs are of very high quality and aesthetics and summarize the results very well. 
In addition, the use of a very sophisticated analysis (multiparametric linear modeling) and the detailed presentation of this analysis, and the R packages used are highly creditable. 
However, several weaknesses, including, biochemically wrong assumptions regarding the assessed parameters, lack of transparency regarding the pre-processing of the data, and measurement comparability between the assessed cohorts limit its significance and impact. 
The quality of the paper also suffers from a lack of references, scientifically inadequate language, and multiple spelling and grammatical errors. Because of the many issues that arise, I would not recommend the manuscript for publication in _Psychoneuroendocrinology_.</span>

We thank the reviewer for the careful assessment of our manuscript and valuable feedback which were very helpful in revising our work. 
First we would like to apologize for imprecise wording, not including enough references and any language errors. 
We have reworked large parts of the manuscript in order to convey our research aims, results and their interpretation in a more clear way. 
For these reasons, we re-formatted the analysis pipeline and the manuscript in the following way:

* We unified nomenclature of the metabolic pathways investigated in our manuscript and term them now 'kynurenine pathway' and 'catecholamine pathway' in line with prominent review articles [@Badawy2017; @Lukic2022; @Meiser2013]

* We reorganized and rewrote many parts of __Introduction__.

* We also more explicitly and clearly state the ‘gap in the literature’ and our research aims in the last paragraph of __Introduction__.

* We included an __Analysis endpoints__ section in __Methods__. Here we describe our analytic approach, define outcome variables and candidate explanatory factors.

* We show the analysis strategy for the SIMMUN and INCOV cohort in __Supplementary Figure S1__. We hope that this makes it easier to follow our analysis. Additionally we restrict the analyzed SIMMUN and INCOV cohorts to participants with the complete set of outcome and explanatory variables (__Supplementary Figure S1__, __Supplementary Table S1__ and __S3__). This makes the results of particular analyses more easily comparable to one another. Additionally, we compared the excluded and analyzed individuals as presented in __Supplementary Table S2__ and __Results/Characteristic of the cohorts__.

@@

* We excluded CRP and IL6 from analyses presented in the revised manuscript. Both  markers were found to be slightly above their reference ranges (0.5 mg/dL and 7 pg/mL) solely in 
`r signif(table(stigma$data %>% filter(patient_id %in% stigma$complete_ids) %>% .$crp_class)[2]/nrow(stigma$analysis_tbl) * 100, 2)`% and 
`r signif(table(stigma$data %>% filter(patient_id %in% stigma$complete_ids) %>% .$il6_class)[2]/nrow(stigma$analysis_tbl) * 100, 2)`% of SIMMUN individuals, respectively. They are hence unlikely to be biologically and clinically relevant predictors of the kynurenine and catecholamine pathway activity. Details on this are now described in __Supplementary Methods__.

* We provided information on all analyzed variables of the SIMMUN (__Supplementary Table S1__) and INCOV cohort (__Supplementary Table S3__) with their format, units and, if applicable, normality-stabilizing transformations.

* In order to improve clarity of __Results__, we have now reordered and reformatted this section. We now present the results in a ‘top - bottom’ approach. In brief, both in the SIMMUN and the INCOV cohort, we employ multi-parameter modeling (linear regression or robust linear regression), to identify key explanatory factors significantly affecting the outcome. Subsequently, we investigate those significant predictors in more detail by ‘univariable’ correlation analysis and statistical hypothesis testing. This analysis strategy has recently been used in two COVID-19-related publications of our group [@Hufner2022; @Hufner2023].

@@

As outlined in __Issue 1__, we re-ordered and re-wrote the entire __Introduction__ section. Following discussions in the core author team, we decided to replace the general term ‘psychoneuroimmunological’ with more direct descriptions of the mechanisms we are referring to i.e. the role of inflammation on IDO activity are the effect of glucocorticoids on TDO activity. We have also changed the title of the manuscript.

## Issue 3

<span custom-style = 'reviewer'>__Introduction:__ Some statements made in the introduction are without reference, very general, and the conclusions weakly justified. 
For instance, TRP, KYN, and KYN/TRP are listed as markers of serotonin availability. 
However, the available evidence for the relationship between increased degradation of TRP to KYN and its effects on serotonin availability is controversial as only 1-2% of tryptophan metabolism goes into the serotonin pathway (https://doi.org/10.3390/metabo12060514). 
In addition, line 60 mentions a depressiogenic and an anxiogenic effect of QUIN, and an anxiety- and depression-amplifying effect of kynurenine and catabolites is stated in line 9 (page 3). 
Should this statement refer to existing cross-sectional findings (e.g., higher QUIN in patients with MDD compared to healthy controls) it is very generalized and not sufficiently justified. 
If the statement refers to evidence that suggests a causal relationship, a reference is urgently needed.</span>

We agree that the role of systemic, i.e. peripheral availability of serotonin as well as its association with kynurenine pathway activity in mental health are controversial, and definitely warrants a critical discussion [@Marx2020; @Brown2021; @Moncrieff2022; @Goldstein1999]. 
We have tried to give a more balanced view on the topic in the revised Introduction and Discussion sections, taking the reviewer´s concerns and suggestions into account. 
On one hand, we find that one of the many challenges is that many mechanistic studies are done in animal models and their significance for humans is not well documented [@Brown2021]. 
On the other hand, human studies are quite equivocal and, while there are influential theoretical manuscripts on the issue, surprisingly few original research has been published so far [@Brown2021].

We now give a more clear, but short introduction to the biochemical pathways in __Introduction__. 
We provide critical discussion of the possible role of these pathways in linking mental and physical health. 
In __Discussion__ and __Limitations__ we comment on the controversy over systemic serotonin availability as well as kynurenine and catecholamine pathway in depression and other mental disorders [@Brown2021; @Marx2020; @Schwarcz2012; @Moncrieff2022; @Goldstein1999].

Additionally, to address the question whether TRP levels are associated with systemic serotonin availability in SARS-CoV-2 infection, we modeled circulating serotonin levels in the INCOV cohort [@Su2022] as a function of its precursor and competitor metabolites (TRP, KYN, QUIN), timepoint of SARS-CoV-2 infection, inflammatory cytokines and age. As presented in Figure 2, we found a significant positive association of plasma TRP and serotonin in multi-parameter modeling, which is expected to be independent of age, inflammation and infection timepoints. In a more detailed correlation and network analysis of the INCOV dataset (Figure 4A, Supplementary Figure S6), we observed significant, weak-to-moderate strength correlation of circulating serotonin and TRP levels in acute and sub-acute infection.

@@

## Issue 4

<span custom-style = 'reviewer'>__Introduction:__ The introduction sentence of the last paragraph in the introduction seems not adequate for the following hypotheses. 
It highlights that low-grade inflammation (as I understand it, in the case of this study IL-6 and CRP concentrations) is associated with mental health. 
However, the presented hypotheses use IL-6 and CRP and mental stress, anxiety, and depressive symptoms as predictors in their analysis (multi-parameter linear modeling) and TRP, KYN, KYN/TRP, PHE, TRY, PHE/TRY as dependent variables.</span>

We are sorry for this unclarity. 
In the revised manuscript, we rephrased the aims, and define outcomes and explanatory factors. 
Please refer to the last paragraph of __Introduction__ and the subsection __Analysis endpoints__ of __Methods__.

@@

## Issue 5

<span custom-style = 'reviewer'>__Methods:__ while the SIMMUN cohort is described in detail, basic information about the INCOV cohort is missing and should be provided.</span>

We included short methodological information of the INCOV cohort in __Methods__ (subsection __Procedures__) and in __Supplementary Methods__ [@Su2020; @Su2022].

@@

## Issue 6

<span custom-style = 'reviewer'>__Methods:__ Before the statistical analysis, several transformations (log2, ln, and square root) were applied. 
However, it is not transparent how each parameter is transformed. I would appreciate it if this could be further specified.</span>

We provide this information in __Methods__ and __Supplementary Methods__. 
Variable format, transformations and, if appropriate, stratification schemes are presented in __Supplementary Tables S1__ and __S3__.

@@

## Issue 7

<span custom-style = 'reviewer'>Methods: Furthermore, when measured within clinical routine laboratory values of CRP and IL-6 are in many cases lower the limits of detection (e.g., 0.3 mg/ml). 
Was this the case in this study? 
If so, it would be useful for the replicability to know how the authors dealt with this issue.</span>

As specified now in __Methods__ (subsection __Procedures__), values beyond or at their detection limits were substituted with the lower or upper detection limit. 
CRP and IL6 were removed from the revised analysis pipeline for the reasons specified in __Issue 1__.

@@

We apologize for the missing references and cite the seminal HADS paper and a report dealing with the clinically relevant HADS cutoffs [@Bjelland2002; @Zigmond1983] in the revised text. 
We also checked consistency of the PSS-4 and HADS psychometric tools employed in our report. 
Since the HADS scales, and in particular, the HADS depression tool were found to be non-tau-equivalent by factors analysis (__Supplementary Figure S2A__), we decided use McDonald's $\omega$ as a consistency measure instead of the more popular Cronbach's $\alpha$ [@Revelle2022; @McDonald1999]. 
The PSS-4 tool demonstrated a good and the HADS scales an excellent consistency, as presented in __Supplementary Figure S2B__.

## Issue 10

<span custom-style = 'reviewer'>__Results:__ First, it would be useful to distinguish between cohorts in the descriptive characterization of cohorts. 
Furthermore, concrete numbers should be given instead of rough descriptions (e.g., "about half" (page 8, line 19)). 
Second, assessing the effect as "dramatic" is not appropriate in a scientific context. 
If at all, existing assessments based on effect sizes can be made using generally accepted definitions (e.g., according to Cohen).</span>

We concur with the reviewer, that a comparison between the cohorts is useful for interpretation of the results. 
We present significant results of the comparison between the SIMMUN and INCOV cohort in __Supplementary Table S5__ and elaborate on them in __Results/Characteristic of the study cohorts__. 
The cohorts differed in age and sex distribution, rates of SARS-CoV-2 infection and infection severity. 
In particular, most INCOV study participants experienced moderate-to-critical COVID-19 and required hospitalization. 
This may explain the stronger effects of inflammation and infection timepoint on inflammatory cytokines and the metabolites of interest, which we discuss in __Discussion__. 
In the revised text, Figures and Tables we also present effect size metrics for statistical hypothesis tests, correlation and modeling. 
We interpret them as proposed in the handbook by Cohen [@Cohen2013] (__Supplementary Methods__).

@@

## Issue 11

<span custom-style = 'reviewer'>__Results:__ In general, numbers (test statistics, p-values) are missing in the results part and should be added. Additionally, effect sizes should be presented.</span>

We present now p values and effect sizes or model $\beta$ with 95% confidence intervals, as appropriate, for the most important findings in Abstract and Results. 
Yet, for the sake of readability, we were not able to provide significance and effect size information for all effects. 
In particular, we do not specify p values and effect sizes for non-significant effects and state instead clearly that the effect missed statistical significance. 
We abstained from referencing test statistics such as $\chi^2$ or $t$, since they depend on the number of complete observations and hence do not deliver reliable information on the effect size. 
Instead we present p values and effect size statistics in plots and tables.

@@

Following your suggestion, we have replaced the term ‘strata’ in the revised text by 'subgroup' or 'subset'.

## Issue 14

<span custom-style = 'reviewer'>__Introduction:__ The use of subheadings would be helpful to facilitate the understanding of the first part of the introduction.</span>

Since we re-wrote and shortened the __Introduction__ section, we did not included subheadings. 
To facilitate interpretation, each topic of __Introduction__ is described in a separate paragraph. 
We now included subheadings in __Discussion__.

@@

For the sake of clarity, we present characteristic of the SIMMUN and INCOV cohorts in separate tables (__Table 1__ and __Table 2__).

## Issue 18

<span custom-style = 'reviewer'>__Results:__ Page 9 lines 24-41; This paragraph should be in the Methods instead of the Results. Furthermore, the rationale is missing: Why was INCOV used as a comparison or additional study to your own data? In addition, information on the measurement /measurement system (for INCOV) should be provided. Were the measuring probes used in SIMMUN and INCOV comparable?</span>

As suggested, we now list the variables of interest in  __Methods__ instead of __Results__. 
Furthermore, we describe the rationale for using the INCOV dataset in the last paragraph of __Introduction__ and the __Analysis endpoints__ subsection of __Methods__. 
Information on measurement of proteins and metabolites is provided in __Methods__ (subsection Procedures) and __Supplementary Methods__. 
Unfortunately, data incompatibilities of the two cohorts hindered us from analyzing them in one comprehensive model with direct validation in the other (canonical training - test setup). 
We list this issue in __Limitations__.

@@

We are sorry for this unclarity. 
In the revised text, we name inflammatory markers (SIMMUN: NEO, NLR, INCOV: IL6, IL10, TNF and IFNG) more explicitly.

## Issue 23

<span custom-style = 'reviewer'>__Discussion:__  The discussion of the study results is rather shallow: Results should be better integrated into existing evidence. Furthermore, there is no discussion of the results of depressive symptoms (HADS).</span>

We restructured and improved __Discussion__. 
Large parts of __Discussion__ were reorganized and rewritten as suggested by reviewer 2 (__Issue 40__). 
We hope that you agree with the structure proposed by the other reviewer which we have now followed. 
We have also included new references to back and critically interpret our results.
The __Limitations__ section was also expanded and improved.

@@

We apologize for the typo, it was corrected to 'BIC'.

## Issue 30

### 2023-06-11 paper/markdown/rebuttal_letter.Rmd

<span custom-style = 'reviewer'>__Supplementary Figure S2:__ It is not clear why fitted generalized adaptive models are presented as graphs, as the associated statistic (Spearman's rho) assesses monotone relationships.</span>

We concur with the reviewer's view, that visualization of trends for monotone associations is problematic. 
Due to profound changes in the structure of the revised manuscript, we do not use scatter plots to illustrate monotone correlations between non-normally distributed variables any more. 
Correlation analysis results in the INCOV cohort are now presented as simple bubble correlograms or visualized as force-directed graphs [@Briatte2021; @Csardi2006].

@@

# Reviewer 2

## Issue 31

<span custom-style = 'reviewer'>In this paper, the authors show the relationship between alterations in markers of inflammation, aminoacid precursors, and symptoms of depression, anxiety, and stress in individuals COVID-19-positive or not. 
They relate the results of these observations to the results obtained in another immune and metabolic study carried out at different time points of the COVID-19 trajectory and already published by a group of different investigators.
The methodology is excellent as can be expected from a research team specialized in the study of inflammation and its impact on the kynurenine metabolism pathway. 
The results are original and sufficiently important to be published. 
The main problem with the current manuscript is the confusion coming from the mode of presentation that mixes up in the results section data from the two studies. 
The discussion is circumvoluted and difficult to follow.</span>

We thank the reviewer for appreciation of out text and the constructive feedback. 
Weaknesses in structure of the analyses were already pointed out by the reviewer 1. 
We addressed them by employing the multi-parameter modeling results for identification of the most vital explanatory factors that affect neurotransmitter metabolism and subsequent more specific correlation analyses and statistical testing with those relevant predictors. 
To make the analysis of the two investigated cohorts easier to follow, we separated the results obtained with the SIMMUN and INCOV cohort. 
In more detail, we employ the SIMMUN cohort as an ‘exploratory’ cohort to identify factors affecting TRP, KYN, KYN/TRP, PHE and PHE/TYR from a wider palette of demographic, clinical, psychometric, inflammatory and SARS-CoV-2 associated variables. Subsequently, we attempted to validate the most important ones, age, inflammation and SARS-CoV-2 infection, in the INCOV cohort. Finally, we re-wrote __Discussion__ with a special focus on the crosstalk between inflammation and neurotransmitter biosynthesis and, in part controversial, relevance of systemic neurotransmitter availability for the central nervous system and mental health.

@@

Yes, in the manuscript  text the term ‘collective’ had the same meaning as ‘cohort’. 
We removed the word 'collective' since this obviously was not easy to understand and use consistently 'cohort' instead. 
For practical reasons, we avoid the wording ‘population sample’, which may suggest a random and hence representative draw from the general population. 
In fact neither the SIMMUN cohort enriched in psychiatric patients nor the INCOV cohort enriched in hospitalized COVID-19 patients pose a representative sample of the general population during the pandemic (see __Limitations__).

## Issue 35

<span custom-style = 'reviewer'>The introduction ends with the statement that the study was conducted in an exploratory manner. Was that the objective of the present study in the absence of any specific hypothesis?</span>

As also pointed out by reviewer 1, the last paragraph of __Introduction__ was not written clearly enough in the initial manuscript version. 
In the final paragraph of __Introduction__, we now present the ‘gap in the literature’ and the specific aim of our analysis. 
Our study did  not have a fully exploratory character, since the candidate explanatory factors: age, sex, body mass index, mental and physical disorders, smoking and alcohol consumption, depression, anxiety, mental stress and inflammation have already been described to be associated with serotonin availability, and kynurenine and catecholamine pathway activity in previous research, also from our own research consortium [@Hufner2019; @Hufner2020; @Hufner2021; @Hufner2015].

@@

We agree and rearranged the complete __Results__ section. 
Please refer to __Issues 1__ and __31__ for details.

## Issue 39

<span custom-style = 'reviewer'> You utilize the terms mental disease, mental stress, mental symptoms, etc. Be more specific and replace these terms by what you measured, i.e., HADS-anxiety, HADS-depression, and PSS. 
You mention somatic symptoms in the discussion but what they refer to remains obscure. 
You also refer to somatic diseases. 
Pathologists would rather speak of physical disorders as opposed to psychiatric disorders.</span>

Thank you for your suggestions. 
We have specified the terms according to your suggestions throughout the revised manuscript. 
We removed the sentence from __Discussion__ referring to somatic symptoms, since those were not systematically surveyed in the SIMMUN cohort. 
Following your suggestion, we use the terms 'mental disorders' and 'physical disorders' consequently in the revised text.

@@

## Issue 40

<span custom-style = 'reviewer'>The discussion is convoluted and needs to be simplified. 
It would be easier to follow if you would first discuss the significance of the association between inflammation and neurotransmitter precursors and then the association between these two sets of factors and what you call mental health measured by HADS and PSS, first in the immune -> HDAS symptomatology or kynurenine -> HADS symptomatology and then in the stress -> immune direction, remembering that in the case of your own cohort, the cross-sectional nature of your study does not allow you to get further than describing associations.</span>

We thank the reviewer for this specific suggestion on the discussion section which we have gratefully taken up and rewritten large parts of __Discussion__ according to the following scheme:

@@

* A brief generalizing summary of the results obtained with both cohorts and discussion of discrepant findings

* Effects of inflammation and infection, including SARS-CoV-2, on the kynurenine pathway as well as systemic levels of serotonin

@@

* Effects of inflammation and infection on the catecholamine pathway and systemic availability of dopamine

* Mutual effects of mental health-related factors (e.g. depression, anxiety and mental stress) on activity of the kynurenine and catecholamine pathways, and systemic serotonin and dopamine availability. We provided a balanced and critical discussion of the evidence here.

@@

# Reviewer 3

## Issue 42

<span custom-style = 'reviewer'>Considering the disease SARS CoV-2 and its consequences in the form of post-covid syndrome, the presented work is highly relevant.</span>

We thank the reviewer for appreciation of our manuscript and the valuable feedback.

@@

## Issue 43

<span custom-style = 'reviewer'>In the abstract, the authors introduce ...markers of serotonin availability (TRP, KYN, KYN/TRP ratio)....(line 24). 
This statement is not accurate, as Serotonin is formed from TRP, but there is no KYN in its catabolic pathway. 
Only if the catabolism of TRP moves more to the kynurenine pathway, less TRP goes to the serotonin pathway, and the availability of serotonin is reduced. 
This argument is similar to the line 44.</span>

We absolutely agree with the reviewer. 
We now consistently use the terms ‘serotonin’, ‘kynurenine pathway’ (TRP, KYN, QUIN) and ‘catecholamine pathway’ (PHE, TYR, dopamine sulfate) in the revised manuscript.

@@

## Issue 44

<span custom-style = 'reviewer'>The principle point is that it would help the reader to orient himself in such a large amount of data if the authors completed the manuscript with a design scheme of the INCOV and SIMMUN study</span>

Thank you for the suggestions. 
As described in __Issues 1__ and __38__, we present the results in cohort-wise manner, for the sake of clarity. 
We also included the analysis strategy scheme as __Supplementary Figure S1__. 
Unfortunately we had already reached the maximum figure number allowed so that this could not be included into the main text.

### 2023-06-11 paper/markdown/issues_07062023.Rmd
Feedback from 08 - 09.06.2023
Changes in headings and labels in Figures, Tables, Supplementary Figures and Supplementary Tables. Changes in the manuscript and supplement text.

---
title: "The effect of inflammation, SARS-CoV-2 infection, age and mental health on serotonin, kynurenine and catecholamine pathway activity"
subtitle: "Issues"
author: "PSY Team"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

>>>>

---
title: "The effect of inflammation, SARS-CoV-2 infection, age and mental health on serotonin, and kynurenine and catecholamine pathway metabolites"
subtitle: "Issues 07.06.2023"
author: "PSY Team"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
